---
title: "Joint modeling of endpoints for RCT"
author: "Lars Mølgaard Saxhaug"
date: "`r Sys.Date()`"
output: github_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstanarm) # devtools::install_github("stan-dev/rstanarm", ref = "feature/survival", build_vignettes = FALSE)
library(tidybayes)
library(rio)
library(here)
```

Recreation of analysis from ["Joint modeling of endpoints can be used to answer various research questions in randomized clinical trials"](https://www.sciencedirect.com/science/article/pii/S0895435622000701#bib33) [@vaneijk2022] using {rstanarm}[@rstanarm]

```{r import}
survival_data <- import(here("data/VPA_SURVIVAL_dataset.xlsx")) |> 
  janitor::clean_names()
str(survival_data)
head(survival_data)

long_data <- import(here("data/VPA_ALSFRS_dataset.xlsx")) |> 
  janitor::clean_names()
str(long_data)
head(long_data)

```

```{r rstanarm_fitting}
# check if file exists, only run model if it does not
if (file.exists(here("output/fits/model1.Rdata"))) { 
  model1 <- import(here("output/fits/model1.Rdata"))
  
} else {
  model1 <-
    stan_jm(
      formulaLong = total ~ avisit + avisit:trt + (avisit + I(avisit ^ 2) | id), # longitudinal mixed effects model
      dataLong = long_data,
      formulaEvent = Surv(stime, status) ~ trt, # survival model 
      dataEvent = survival_data,
      time_var = "avisit",
      priorEvent = rstanarm::normal(0, 1), # prior on treatment effect in survival model
      iter = 4000,
      adapt_delta = 0.99
    )
  save(model1, file = here("output/fits/model1.Rdata"))
  
}

```

```{r piror summary}
prior_summary(model1) # 
```

```{r summary of model}
summary(model1)
```
#### Diagnostics
```{r}
bayesplot::mcmc_trace(model1,regex_pars = "Sigma.*")
bayesplot::mcmc_nuts_divergence(bayesplot::nuts_params(model1),bayesplot::log_posterior(model1))

```



#### Posterior predictive check of survival

```{r ps_check}
ps_check(model1)

```

#### Plot of survival probabilities conditional on treatment

Averaging across individual-specific predictions within each treatment group, derived from

<https://discourse.mc-stan.org/t/survival-predictions-by-categorical-covariates-in-stan-jm/9516/4>

```{r surv_plot_data, cache=TRUE}
id0 <- survival_data$id[survival_data$trt==0]
id1 <- survival_data$id[survival_data$trt==1]
p0 <- posterior_survfit(model1, ids = id0, standardise = TRUE, times = 0 )
p1 <- posterior_survfit(model1, ids = id1, standardise = TRUE, times = 0 )

p0 |> as_tibble() |> 
  mutate(trt=0) |> 
  bind_rows(p1 |> 
              mutate(trt=1)) |> 
  ggplot(aes(x=avisit,y=median,fill=factor(trt),colour=factor(trt)))+
  geom_line()+
  geom_ribbon(aes(ymin=ci_lb,ymax=ci_ub),alpha=0.4,colour=NA)+
  scale_y_continuous(name = "Survival")+
  scale_colour_manual(values = c("pink","steelblue"))+
  scale_fill_manual(values = c("pink","steelblue"))+
  labs(fill="Treatment group",colour="Treatment group")+
  theme_tidybayes()

```

#### Bibliography
